Panarchy: ripples of a boundary concept

### 2022-03-07 paper/paper_panarchy.Rmd
implementing suggestions from reviewers and co-authors

---
title: "Panarchy: ripples of a boundary concept"
author: |
    | \small Juan Rocha^1,2,3^, Linda Luvuno^4^, Jesse Rieb^5^, Erin Crockett^6^,  Katja Malmborg^1^, Michael Schoon^7^, Garry Peterson^1^
    | \small
    | \footnotesize ^1^Stockholm Resilience Centre, Stockholm University, Kräftriket 2B, 10691 Stockholm
    | \footnotesize ^2^Future Earth, The Swedish Royal Academy of Sciences, Lilla Frescativägen 4A, 104 05 Stockholm
    | \footnotesize ^3^South American Institute for Resilience and Sustainability Studies, Maldonado, Uruguay
    | \footnotesize ^4^Center for Complex Systems in Transition, Stellenbosch University, South Africa
    | \footnotesize ^5^Department of Geography, McGill University, 805 rue Sherbrooke ouest, Montréal (Québec) H3A 0B9, Canada
    | \footnotesize ^6^Department of Natural Resource Sciences, McGill University, 21111 Lakeshore Road, 
    | \footnotesize Sainte-11 Anne-de-Bellevue, QC, H9X 3V9, Canada
    | \footnotesize ^7^School of Sustainability, Arizona State University, Tempe, Arizona, USA
abstract: |
  How do social-ecological systems change over time? In 2002 Holling and colleagues proposed the concept of Panarchy, which presented social-ecological systems as an interacting set of adaptive cycles, each of which is produced by the dynamic tensions between novelty and efficiency at multiple scales. Initially introduced as a conceptual framework and set of metaphors, panarchy has gained the attention of scholars across many disciplines, and its ideas continue to inspire further conceptual developments. Almost twenty years after this concept was introduced we review how it has been used, tested, extended and revised.  We do this by combining qualitative methods and machine learning. Document analysis was used to code panarchy features that are commonly used in the scientific literature (N = 42), a qualitative analysis that was complemented with topic modeling of 2177 documents. We find that the adaptive cycle is the feature of panarchy that has attracted the most attention. Challenges remain in empirically grounding the metaphor, but recent theoretical and empirical work offer some avenues for future research.
bibliography: [panarchy.bib, packages.bib]
csl: "/Users/juanrocha/Documents/styles/ecology.csl"
output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
  rtf_document:
    toc: no
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt
linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, fontspec, array, tabularx}
- \setmainfont{Arial}
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}
---

@@

Almost two decades ago, the edited book "Panarchy: Understanding Transformations
in Human and Natural Systems" [@Gunderson:2002vk], presented a synthetic
perspective of how a group of social-ecological researchers associated with the
Resilience Alliance understood change in social-ecological systems. The concept
of Panarchy was a key focus of this influential book. In the late 1990s and
early 2000s, the Resilience Alliance was a productive, innovative, and highly
collaborative group of interdisciplinary scientists that focused on addressing
social-ecological problems. They did this by combining insights from the social
and natural sciences, arts, and humanities, as well as by bridging theory and
practice [@Parker:2012eo].

The Panarchy concept builds on Holling's adaptive cycle [@biosphere:hJ4c_Ua8]
by extending the idea across spatial and temporal scales. It remains a boundary
object that has inspired research topics, enabled collaborations and nurtured
new scientific frameworks. The ideas put forward have been applied in field
studies, archaeology, mathematical models, participatory work, and scenario
development. Panarchy has inspired resilience assessments and guided decision
making. In this article we pay tribute to the book by studying how the concepts
and metaphors proposed have been further developed in the academic literature.
We also document criticism of the concepts and identify key research frontiers.

Panarchy proposes that it is useful to conceptualize systems in terms of
interacting adaptive cycles. The adaptive cycle was an idea first proposed by
Holling based upon his experience working and studying managed ecosystems
[@biosphere:hJ4c_Ua8]. It was meant to be a conceptual tool that focused
attention on processes of destruction and reorganization, which had been
neglected in comparison to those of growth and conservation.

@@

Panarchy is not a theory of what it is, but a metaphor of what might be
[@Gunderson:2002vk]. It is not a predictive tool, but aims to understand
adaptive change. The adaptive cycle is hypothesized to exist within a
three-dimensional space defined by three properties: potential, connectedness,
and resilience [@Gunderson:2002vk]. Potential refers to the capital available
to the system, for example nutrients and carbon captured by a forest, or human
capital - skills and knowledge accumulated to run the economy. Connectedness is
a proxy of the structure of the system; it is the network of interactions and
strength between its elements. Resilience is the capacity of any system to
absorb disturbance and reorganize while undergoing change so as to retain
essentially the same identity, as determined by its function, structure, and
feedbacks [@Anonymous:2016fv].

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 2.5in]{fig1_panarchy}
\caption{\textbf{Panarchy} is an heuristic of nested adaptive cycles that serves to represent a variety of systems and environmental problems. Adaptive cycles (A) at different scales of the hierarchy (B) can be connected through remember and revolt cross-scale interactions.}
\label{fig:fig1}
\end{figure*}
```
Panarchy posits that systems are organized in nested hierarchies across space
and time, where each level of the hierarchy is a subsystem that can be in a
different phase of the adaptive cycle. These phases are inspired by mathematical
models used in economics and ecology, but broadly describe patterns of growth,
collapse, and recovery that are common to populations, ecological communities,
markets, or political organizations. While each subsystem in the hierarchy can
be at a different phase of the adaptive cycle, these subsystems can influence
each other through cross-scale interactions called *"revolt"* and *"remember"*
[@Gunderson:2002vk].

The three-dimensional space has corners with attractors that can cause the
adaptive cycle to stagnate: poverty and rigidity traps [@Gunderson:2002vk].
Poverty traps are described in panarchy as maladaptive states where potential,
connectedness and resilience are low. Poverty traps are a series of feedback
mechanisms that reinforce impoverished states [@Bowles:2006hx; @Maru:2012dm;
@Allison:2004te], limiting the system's capacity to innovate and increase
potential. The opposite corner, where potential, connectivity and resilience are
high, is another maladaptive space called a rigidity trap. In that corner there
is little space for experimentation and innovation. Examples include systems
where ecological resilience has been extensively replaced by artificial
processes to maintain the system such as levees, flood barriers, or chemical
control of pests [@Holling:1996wx].

Panarchy offers a rich conceptual framework for understanding environmental
problems. While inspired by several mathematical constructs (e.g. cycles, traps,
scaling laws), it is general enough to invite scholars from multiple
disciplinary backgrounds, ontologies and epistemologies, to collaborate around
research questions and applied problems. As such, it is useful as a boundary
object and can be used empirically or metaphorically. The book itself presents a
series of case studies where geographers, economists, political scientists, and
ecologists have demonstrated the utility of the framework to their area of
research.

However, the extent to which these concepts have gained empirical support beyond
the metaphor, or how they have evolved over time, remain open questions. For
example, the adaptive cycle is a useful metaphor to look back at history and
organize events and periods around phases, but it remains challenging to
identify the phase of a system in real time, or to use panarchy to project
future trajectories. Scientific theories are bodies of knowledge that answer the
question why? The answer implies causal reasoning. Regardless of discipline, a
theory should therefore provide a series of testable expectations that help
distinguishing the situations when it is a good explanation of reality, versus
situations when it falls short. The practice of science implies avoiding
confirmation bias and testing when explanations are valid representations of
reality. For panarchy to become a useful theory in sustainability science, it
needs to stand the rigor of testing, and being able to explain phenomena in
social-ecological systems across different contexts. We identify progress and
knowledge gaps in the scientific practice inspired by panarchy.

Here we review the academic literature of the last two decades to trace how the
ideas proposed in Panarchy have evolved, to what type of problems they have been
applied and found useful, and finally, what key research frontiers remain.

@@

*Data*: We used the Web of Science, Scopus, and GoogleScholar to survey academic
literature that has used or referenced works that trace back to Panarchy
[@Gunderson:2002vk]. We extracted complete records from the Scopus database
that matched the search for *"panarchy OR adaptive cycle"* (N = 595), or the
search *"(panarchy OR adaptive cycle) AND resilience"* (N = 278). The data was
combined with all papers (N = 1923) that cited the inaugural paper that
introduced the book to the academic community [@Holling:2001cu]. Records with
missing abstracts were dropped (N = 191), and records with missing year were set
to 2020 given that they are accepted manuscripts with digital object identifier
(doi) scheduled to be published later in 2020 or 2021. Our final sample is
`r dim(dtm)[1]` documents.

@@

Based on the reading of Panarchy, we first developed a series of categorical
variables that captured the main components of the scientific framework. The
variables were presented and received feedback at the science meeting of the
Resilience Alliance in 2019. We used the variables to annotate qualitative
aspects such as use of the adaptive cycle, identification of its phases, whether
the paper is conceptual, modeling or an empirical study. When empirical, we
recorded the temporal and spatial scales of the case study. We also
distinguished empirical cases from empirical constructs. An empirical case is
when a study is trying to test an idea or proposition put forward by Panarchy
(e.g. a cross-scale interaction, or that the back loop is faster than the forth
loop). An empirical construct is when Panarchy is used as inspiration to
interpret a case study, typically with a historical component to it (e.g.
identify the phases of adaptive cycle); but it does not find support for or
against a Panarchy preposition. That is, there was no mechanism in place to
avoid confirmation bias.

We also identified what aspects of panarchy were most used in the papers. For
example whether there is emphasis on cross-scale interactions, or poverty and
rigidity traps. We used text annotations to capture potential criticisms as well
as the methods used. We did a preliminary scoping and coding of papers, and
through an iterative process we kept track of potential fields under review by
our corpus, or simply absent of it. We coded papers until we reached conceptual
saturation, that is, additional papers were not proving additional categories of
the already coded. The code book is available at
[doi.org/10.6084/m9.figshare.13490919](http://doi.org/10.6084/m9.figshare.13490919).

@@

Panarchy, the book published in 2002, has been cited over 7200 times according
to Google Scholar. The scientific paper that introduced the book to the
scientific community [@Holling:2001cu] has received, at the time of writing,
1715 citations in the Web of Science, and 1923 in Scopus. Roughly half of the
citations have come from environmental (28%) and social sciences (22%). Computer
science (2.1%) and arts and humanities (2.6%) have been the least represented.

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 5in, height = 4in]{fig2_topicnumber}
\caption{\textbf{Algorithm and number of topic selection} Gibbs sampling maximizes entropy and log-likelihood estimation, making it a suitable algorithm for our data (A). Increasing the number of topics (from 5-100) shows that $\alpha$ decreases, suggesting that despite the larger number of topics, a few of them suffice to describe most papers (B). Log-likelihood is maximized for 50 topics followed closely with 25 and 100. While 50 topics is marginally better. We choose 25 because the $\alpha$ parameter suggest preferring lower numbers, or a more parsimonious solution to avoid over fitting. Note that perplexity cannot be calculated for Gibbs sampling, hence the missing value in A and absence in B.}
\label{fig:algorithm}
\end{figure*}
```
Comparing algorithm performance revealed that Gibbs sampling was the best fit
for the data (Fig \ref{fig:algorithm}). As a rule of thumb, an ideal method
should maximize entropy and the log-likelihood estimation while minimizing
perplexity [@Grun:2011tb]. Gibbs sampling maximised both entropy and
likelihood with our data when compared to other alternatives. The second best
was the variational-expectation maximisation algorithm (VEM) when $\alpha$ was
not set constant. $\alpha$ is a hyper parameter that weights the evenness of
topic distribution. A lower $\alpha$ than default values indicates that the
documents can be described by rather fewer topics, or that its distribution is
very uneven. In fact, we observed that increasing the number of topics from 5 to
100 topics did increase entropy at expense of reducing $\alpha$, meaning that
despite the larger number of topics, the main content of a document is still
captured by a few of them. The log-likelihood maximization stopped at around
25-50 topics. Thus, we restricted the rest of our analysis to 25 topics.

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 5in, height = 5in]{fig3_topics25}
\caption{\textbf{Panarchy topics over time} The number of papers per year (A) with a maximum of 324 in 2019. The proportion of papers per year (B) and the relative proportion of topic content per year (C) do not show strong trends for the time window with most papers (gray area in A). In (B), a reference dotted horizontal line marks the level where all topics are equally dominant. Each topic is summarized in (D) by the top 10-words that best describe them according to the posterior probability of our model fit. Fig \ref{fig:figS1} expands (B) presenting time series for each topic. All time series exclude 2021 to account for the artificial decline of papers at the end of the time series as shown in A.}
\label{fig:topics25}
\end{figure*}
```
A topic is a set of words that are ranked according to a probability that they
represent an underlying content of the document (Fig \ref{fig:topics25}). For
example, the words "resilience", "adaptive", or "capacity" have a high
probability to capture the content of topic 14. Papers early in our time series
(2001-3) have high content largely dominated by topics 14 on resilience, 9 on
resource management, and 23 on cross-scale interactions. Towards the end of the
time series, topic 7 on urban systems and 25 on local communities and knowledge
have spikes up to 9% of the content of each year (2019-2020). For comparison, if
all topics were equally represented in the content, they would have 4% in the
corpus (grey line in Fig \ref{fig:topics25}B). The topics presented lack names
for reproducibility purposes. While here we present e.g. topic 7 as the urban
topic, if a researcher replicates the analysis that person will find the same
list of words and probability distributions listed under a topic with a
different number. That is because the labeling of topics is random and we do not
want to bias the reader to over interpret the word patterns found.

@@

The human-coded document analysis revealed that the most common feature of
panarchy in the literature was the adaptive cycle (81%, N = `r nrow(qual)`)
followed by cross-scale interactions (Fig \ref{fig:qual}). Poverty and rigidity
traps were less studied features in our sample (14% and 19% respectively), even
when accounting for slightly different terminology such as "lock-ins". 52.4% of
the papers analyzed were conceptual papers, many without defined methods or
discernible research question. Roughly half (47%) of the papers were conceptual
or only used panarchy as a metaphor. Over 40% of papers in our sample went a
step further and used panarchy as an empirical construct, for example by
attempting to identify the phases of the adaptive cycle (61%). Out of the 22
empirical cases, 6 were at the time scale of centuries, 7 on decades, 6 on
years, and 1 in weeks. Spatially, 4 were at a city scale, 17 regional, and 1
national. Empirical papers predominantly used qualitative methods (77%, N = 22)
and were generally retrospective historical reviews.

Several of the papers reviewed (81%) attempted to identify the adaptive cycle
either through qualitative or quantitative methods. Identification of adaptive
cycles has found applications in a wide range of disciplines and research
problems, from delimitation of periods in archaeology and anthropology
[@Redman:2003bd], systemic interactions between forest-related sectors in Sweden [@moen2010interlocking], financial crises in Europe [@Castell:2020dg], to
traffic jams in China [@Zeng:2020hu]. The adaptive cycle has been used as a model to structure rich empirical information about participation in adaptive management and, based on that analysis, to develop recommendations for more flexible stakeholder engagement [@stringer2006unpacking]. The adaptive cycle has also been used to provide a more mechanistic description of the history of a variety of specific social-ecological systems, including the Galapagos Islands [@gonzalez2008rethinking], the Western Australian Agricultural Region [@allison2004resilience], the Erhai Lake watershed in southwest China [@dearing2008landscape], and the cities of Cambridge and Swansea in the UK [@simmie2010economic]. The adaptive cycle has also been used to describe more general social and ecological processes, such as lake eutrophication [@carpenter2001metaphor] and the development of regional industrial clusters [@martin2011conceptualizing].

Recent empirical tests of the
adaptive cycle used innovative types of data and methods. For example,
information transfer methods based on entropy have been suggested to approximate
relevant components of a system and the empirical proxies of the axes within
which the adaptive cycle unfolds: potential, connectedness, and resilience
[@Castell:2020dg]. Percolation methods combined with big data have been proven
useful for testing hypotheses of regime shifts in urban systems (originally
proposed in Panarchy [@Gunderson:2002vk]), and deriving the temporal and
spatial scales at which the adaptive cycle emerges [@Zeng:2020hu].

@@

Given the strong predominance of conceptual papers, it is not surprising that
one of the key frontiers of research is to empirically ground such
conceptualizations in a way that one can test theoretical expectations beyond
metaphors. Metaphors have played an important role as boundary objects enabling
interdisciplinary dialogues. We observed this on the wide variety of
applications that panarchy has inspired, in fields from anthropology to
engineering. In the empirical realm, the adaptive cycle is the dimension of
panarchy with which researchers engage the most, but it is dominated by
qualitative and retrospective studies. That means, we still lack the theory and
methods to gather observations and identify a system's current adaptive cycle
phase, or the probability of it transitioning to a new phase.

The high level of conceptualization but lack of empirically grounded testing of
panarchy theory came across as one of the major limitations, given our sample of
papers reviewed (N=`r nrow(qual)`). What makes a theory a scientific theory is
that the plurality of scientific approaches (e.g. positivism, constructivism)
should have mechanisms in place to avoid confirmation bias, allowing the
emergence of alternative explanations, and identify the circumstances under
which a theory may fail [@Hansson:2017tr]. Many of the papers we read
confirmed but did not challenge the expectations that Panarchy put forward,
though a few of the most recent works did falsify some of the ideas proposed
--e.g. the back loop is faster than the front one [@Castell:2020dg]. We
believe more work is needed to better understand the circumstances where
Panarchy works, and when it falls short. In other words, to refine the
scientific framework into a scientific theory.

>>>>

---
title: "Panarchy: ripples of a boundary concept"
author: |
    | \small Juan C. Rocha^1,2,3^, Linda Luvuno^4^, Jesse Rieb^5^, Erin Crockett^6^,  Katja Malmborg^1^, Michael Schoon^7^, Garry D. Peterson^1^
    | \small
    | \footnotesize ^1^Stockholm Resilience Centre, Stockholm University, Kräftriket 2B, 10691 Stockholm
    | \footnotesize ^2^Future Earth, The Swedish Royal Academy of Sciences, Lilla Frescativägen 4A, 104 05 Stockholm
    | \footnotesize ^3^South American Institute for Resilience and Sustainability Studies, Maldonado, Uruguay
    | \footnotesize ^4^Center for Complex Systems in Transition, Stellenbosch University, South Africa
    | \footnotesize ^5^Department of Geography, McGill University, 805 rue Sherbrooke ouest, Montréal (Québec) H3A 0B9, Canada
    | \footnotesize ^6^Department of Natural Resource Sciences, McGill University, 21111 Lakeshore Road, 
    | \footnotesize Sainte-11 Anne-de-Bellevue, QC, H9X 3V9, Canada
    | \footnotesize ^7^School of Sustainability, Arizona State University, Tempe, Arizona, USA
abstract: |
  How do social-ecological systems change over time? In 2002 Holling and colleagues proposed the concept of panarchy, which presented social-ecological systems as an interacting set of adaptive cycles, each of which is produced by the dynamic tensions between novelty and efficiency at multiple scales. Initially introduced as a conceptual framework and set of metaphors, panarchy has gained the attention of scholars across many disciplines, and its ideas continue to inspire further conceptual developments. Almost twenty years after this concept was introduced we review how it has been used, tested, extended and revised.  We do this by combining qualitative methods and machine learning. Document analysis was used to code panarchy features that are commonly used in the scientific literature (N = 42), a qualitative analysis that was complemented with topic modeling of 2177 documents. We find that the adaptive cycle is the feature of panarchy that has attracted the most attention. Challenges remain in empirically grounding the metaphor, but recent theoretical and empirical work offer some avenues for future research.
bibliography: [panarchy.bib, packages.bib]
csl: "/Users/juanrocha/Documents/styles/ecology.csl"
output:
  rtf_document:
    toc: no
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt
linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, fontspec, array, tabularx}
- \setmainfont{Arial}
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}
---

@@

Almost two decades ago, the edited book "Panarchy: Understanding Transformations
in Human and Natural Systems" [@Gunderson:2002vk], presented a synthetic
perspective of how a group of social-ecological researchers associated with the
Resilience Alliance understood change in social-ecological systems. The concept
of panarchy was a key focus of this influential book. In the late 1990s and
early 2000s, the Resilience Alliance was a productive, innovative, and highly
collaborative group of interdisciplinary scientists that focused on addressing
social-ecological problems. They did this by combining insights from the social
and natural sciences, arts, and humanities, as well as by bridging theory and
practice [@Parker:2012eo].Panarchy remains a boundary
object that has inspired research topics, enabled collaborations and nurtured
new scientific frameworks [@Parker:2012eo]. The ideas put forward have been applied in field
studies, archaeology, mathematical models, participatory work, and scenario
development [@gunderson2022applied]. Panarchy has inspired resilience assessments and guided decision
making. In this article we pay tribute to the book by studying how the concepts
and metaphors proposed have been further developed in the academic literature.
We also document criticism of the concepts and identify key research frontiers.

The panarchy concept builds on Holling's adaptive cycle [@biosphere:hJ4c_Ua8]
by extending the idea across spatial and temporal scales. Panarchy proposes that it is useful to conceptualize systems in terms of
interacting adaptive cycles. The adaptive cycle was an idea first proposed by
Holling based upon his experience working and studying managed ecosystems
[@biosphere:hJ4c_Ua8]. It was meant to be a conceptual tool that focused
attention on processes of destruction and reorganization, which had been
neglected in comparison to those of growth and conservation.

@@

"_Panarchy is not a theory of what it is, but a metaphor of what might be_"
[@Gunderson:2002vk 32]. It is not a predictive tool, but aims to understand
adaptive change. The adaptive cycle is proposed to exist within a
three-dimensional space defined by three properties: potential, connectedness,
and resilience [@Gunderson:2002vk]. Potential refers to the capital available
to the system, for example nutrients and carbon captured by a forest, or human
capital - skills and knowledge accumulated to run the economy. Connectedness is
a proxy of the structure of the system; it is the network of interactions and
strength between its elements. Resilience is the capacity of any system to
absorb disturbance and reorganize while undergoing change so as to retain
essentially the same identity, as determined by its function, structure, and
feedbacks [@Anonymous:2016fv].

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 2in]{fig1_panarchy}
\caption{\textbf{Panarchy} is an heuristic of nested adaptive cycles that serves to represent a variety of systems and environmental problems. Adaptive cycles (A) at different scales of the hierarchy (B) can be connected through remember and revolt cross-scale interactions. (C) relates the adaptive cycle to the potential, connectedness and resilience axis, reproduced from Panarchy edited by Lance Gunderson and C.S. Holling, copyright \textcopyright 2002 Island Press. Reproduced with permission of Island Press, Washington, DC.}
\label{fig:fig1}
\end{figure*}
```
Panarchy posits that systems are organized in nested hierarchies across space
and time, where each level of the hierarchy is a subsystem that can be in a
different phase of the adaptive cycle. These phases are inspired by mathematical
models used in economics and ecology, but broadly describe patterns of growth,
collapse, and recovery that are common to populations, ecological communities,
markets, or political organizations. While each subsystem in the hierarchy can be at a different phase of the adaptive cycle, these subsystems can particularly influence each other through cross-scale interactions called *“revolt”*, where release in a smaller, faster sub-system can trigger release in a larger, slower sub-system,  and *“remember”*, in which the structure of a larger-slower sub-system shapes the dynamics of reorganization of a smaller, faster sub-system
[@Gunderson:2002vk].

The three-dimensional space defined by potential, connectedness and resilience, has corners with attractors that can derail the
adaptive cycle: poverty and rigidity traps [@Gunderson:2002vk].
Poverty traps are described in panarchy as maladaptive states where potential,
connectedness and resilience are low. Poverty traps are a series of feedback
mechanisms that reinforce impoverished states [@Bowles:2006hx; @Maru:2012dm;
@Allison:2004te], limiting the system's capacity to innovate and increase
potential. The opposite corner, where potential, connectivity and resilience are
high, is another maladaptive space called a rigidity trap. In that corner there
is little space for experimentation and innovation. Examples include systems
where ecological resilience has been extensively replaced by artificial
processes to maintain the system such as levees, flood barriers, or chemical
control of pests [@Holling:1996wx].

Panarchy offers a rich conceptual framework for interpreting environmental
problems. While inspired by several mathematical constructs (e.g. cycles, traps,
scaling laws), it is general enough to invite scholars from multiple
disciplinary backgrounds, ontologies and epistemologies, to collaborate around
research questions and applied problems. As such, it is useful as a boundary
object and can be used empirically or metaphorically. The book presents a
series of case studies where geographers, economists, political scientists, and
ecologists have demonstrated the utility of the framework to their area of
research.

Scientific theories are bodies of knowledge that answer the
question why? The answer implies causal reasoning. Regardless of discipline, a
theory should therefore provide a series of testable expectations that help
distinguishing the situations when it is a good explanation of reality, versus
situations when it falls short. The practice of science implies avoiding
confirmation bias and testing when explanations are valid representations of
reality. In this paper we explore how the conceptual framework of panarchy has been applied, developed and tested over the past twenty years. For panarchy to become a useful theory in sustainability science, it 
needs to be able to explain phenomena in
social-ecological systems across different contexts, in a way tah tis testable, and withstands the rigour of those tests. Here we review the academic literature of the last two decades to trace how the
ideas proposed in panarchy have evolved. We conclude by suggesting research frontiers for future work.

@@

*Data*: We used the Web of Science, Scopus, and GoogleScholar to survey academic
literature that has used or referenced works that trace back to panarchy
[@Gunderson:2002vk]. We extracted complete records from the Scopus database
that matched the search for *"panarchy OR adaptive cycle"* (N = 595), or the
search *"(panarchy OR adaptive cycle) AND resilience"* (N = 278). The data was
combined with all papers (N = 1923) that cited the inaugural paper that
introduced the book to the academic community [@Holling:2001cu]. Records with
missing abstracts were dropped (N = 191), and records with missing year were set
to 2020 given that they are accepted manuscripts with digital object identifier
(doi) scheduled to be published later in 2020 or 2021. Our final sample is
`r dim(dtm)[1]` documents.

@@

Based on the reading of panarchy, we first developed a series of categorical
variables that captured the main components of the scientific framework. The
variables were presented and received feedback at the science meeting of the
Resilience Alliance in 2019. We used the variables to annotate qualitative
aspects such as use of the adaptive cycle, identification of its phases, whether
the paper is conceptual, modeling or an empirical study. When empirical, we
recorded the temporal and spatial scales of the case study. We also
distinguished empirical cases from empirical constructs. An empirical case is
when a study is trying to test an idea or proposition put forward by panarchy
(e.g. a cross-scale interaction, or that the back loop is faster than the forth
loop). An empirical construct is when panarchy is used as inspiration to
interpret a case study, typically with a historical component to it (e.g.
identify the phases of adaptive cycle); but it does not find support for or
against a panarchy preposition. That is, there was no mechanism in place to
avoid confirmation bias.

We also identified what aspects of panarchy were most used in the papers. For
example whether there is emphasis on cross-scale interactions, or poverty and
rigidity traps. We used text annotations to capture potential criticisms as well
as the methods used. We did a preliminary scoping and coding of papers, and
through an iterative process we kept track of potential fields under review by
our corpus, or simply absent from it. We coded papers until we reached conceptual
saturation, that is, additional papers were not proving additional categories. The code book is available at
[doi.org/10.6084/m9.figshare.13490919](http://doi.org/10.6084/m9.figshare.13490919).

@@

Panarchy, the book published in 2002, has been cited over 8600 times according
to Google Scholar (by March 2022). The scientific paper that introduced the book to the
scientific community [@Holling:2001cu] has received, at the time of writing,
1853 citations in the Web of Science, and 2197 in Scopus. Roughly half of the
citations have come from environmental (28%) and social sciences (22%). Computer
science (2.1%) and arts and humanities (2.6%) have been the least represented.

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 5in, height = 4in]{fig2_topicnumber}
\caption{\textbf{Algorithm and number of topic selection} Gibbs sampling maximizes entropy and log-likelihood estimation, making it a suitable algorithm for our data (A). Increasing the number of topics (from 5-100) shows that $\alpha$ decreases, suggesting that despite the larger number of topics, a few of them suffice to describe most papers (B). Log-likelihood is maximized for 50 topics followed closely with 25 and 100. While 50 topics is marginally better. We choose 25 because the $\alpha$ parameter suggest preferring lower numbers, or a more parsimonious solution to avoid over fitting. Note that perplexity cannot be calculated for Gibbs sampling, hence the missing value in A and absence in B.}
\label{fig:algorithm}
\end{figure*}
```

Topic modelling was done using Gibbs sampling. In our comparison of alternative Latent Dirichlet Allocation algorithms Gibbs sampling provided the best fit
for the data (Fig \ref{fig:algorithm}). As a rule of thumb, an ideal method
should maximize entropy and the log-likelihood estimation while minimizing
perplexity [@Grun:2011tb]. Gibbs sampling maximised both entropy and
likelihood with our data when compared to other alternatives. The second best
was the variational-expectation maximisation algorithm (VEM) when $\alpha$ was
not set constant. $\alpha$ is a hyper parameter that weights the evenness of
topic distribution. A lower $\alpha$ than default values indicates that the
documents can be described by rather fewer topics, or that its distribution is
very uneven. In fact, we observed that increasing the number of topics from 5 to
100 topics did increase entropy at expense of reducing $\alpha$, meaning that
despite the larger number of topics, the main content of a document is still
captured by a few of them. The log-likelihood maximization stopped at around
25-50 topics. Thus, we restricted the rest of our analysis to 25 topics.

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 5in, height = 5in]{fig3_topics25}
\caption{\textbf{Panarchy topics over time} The number of papers per year (A) with a maximum of 324 in 2019. The proportion of papers per year (B) and the relative proportion of topic content per year (C) do not show strong trends for the time window with most papers (gray area in A). In (B), a reference dotted horizontal line marks the level where all topics are equally dominant. Each topic is summarized in (D) by the top 10-words that best describe them according to the posterior probability of our model fit. Fig \ref{fig:figS1} expands (B) presenting time series for each topic. All time series exclude 2021 to account for the artificial decline of papers at the end of the time series as shown in A.}
\label{fig:topics25}
\end{figure*}
```
A topic is a set of words that are ranked according to a probability that they
represent an underlying content of the document (Fig \ref{fig:topics25}). For
example, the words "resilience", "adaptive", or "capacity" have a high
probability to capture the content of topic 14. Papers early in our time series
(2001-3) have high content largely dominated by topics 14 on resilience, 9 on
resource management, and 23 on cross-scale interactions. Towards the end of the
time series, topic 7 on urban systems and 25 on local communities and knowledge
have spikes up to 9% of the content of each year (2019-2020). For comparison, if
all topics were equally represented in the content, they would have 4% in the
corpus (grey line in Fig \ref{fig:topics25}B). The topics presented lack names
for reproducibility purposes. While here we present e.g. topic 7 as the urban
topic, if a researcher replicates the analysis that person will find the same
list of words and probability distributions listed under a topic with a
different number. That is because the labeling of topics is random and we urge readers not to over interpret the word patterns found.

@@

The human-coded document analysis revealed that the most common feature of
panarchy in the literature was the adaptive cycle (81%, 34 out of N = `r nrow(qual)`)
followed by cross-scale interactions (Fig \ref{fig:qual}). Poverty and rigidity
traps were less studied features in our sample (14% and 19% respectively), even
when accounting for slightly different terminology such as "lock-in". About half (52.4% or 22/42) of
the papers analyzed were conceptual papers, many without defined methods or
a clear research question. Roughly half (20/42) of the papers used panarchy as a metaphor. About 40% of papers in our sample (17/42) went a
step further and used panarchy as an empirical construct, for example by
attempting to identify the phases of the adaptive cycle (26/42). Off the 22
empirical cases, 6 were at the time scale of centuries, 7 on decades, 6 on
years, and 1 in weeks. Spatially, 4 were at a city scale, 17 regional, and 1
national. Empirical papers predominantly used qualitative methods (17)
and were primarily retrospective historical reviews.

Most of the papers reviewed (81%, 34 of N = 42) identified adaptive cycles
either through scientific methods, or as analogy or framework to make sense of a narrative. Identification of adaptive
cycles has found applications in a wide range of disciplines and research
problems, from delimitation of periods in archaeology and anthropology
[@Redman:2003bd], systemic interactions between forest-related sectors in Sweden [@moen2010interlocking], financial crises in Europe [@Castell:2020dg], to
traffic jams in China [@Zeng:2020hu]. The adaptive cycle has been used as a model to structure rich empirical information about participation in adaptive management and, based on that analysis, to develop recommendations for more flexible stakeholder engagement [@stringer2006unpacking]. The adaptive cycle has also been used to provide a more mechanistic description of the history of a variety of specific social-ecological systems, including the Galapagos Islands [@gonzalez2008rethinking], the Western Australian Agricultural Region [@allison2004resilience], the Erhai Lake watershed in southwest China [@dearing2008landscape], and the cities of Cambridge and Swansea in the UK [@simmie2010economic]. The adaptive cycle has also been used to describe more general social and ecological processes, such as lake eutrophication [@carpenter2001metaphor] and the development of regional industrial clusters [@martin2011conceptualizing].

Recent empirical tests of the
adaptive cycle used innovative types of data and methods. For example,
information transfer methods based on entropy have been suggested to approximate
relevant components of a system and the empirical proxies of the axes within
which the adaptive cycle unfolds: potential, connectedness, and resilience
[@Castell:2020dg]. Percolation methods combined with big data have been proven
useful for testing hypotheses of regime shifts in urban systems, an idea originally
proposed in panarchy [@Gunderson:2002vk]; and deriving the temporal and
spatial scales at which the adaptive cycle emerges [@Zeng:2020hu].

@@

We examined how the conceptual framework of panarchy has been developed and tested over the past twenty years. We did this using a topic modelling to analyze a large number of papers and a focused review of a selection of panarchy papers.  We find that panarchy has been used an interpretative tool across a wide variety of topics, but has there has been little effort to substantially operationalize or test panarchy.

Sustainability science is defined by the problems it addresses rather than the
disciplines it employs [@Anonymous:2016fv]. This goal of this paper was to explore how the ideas proposed in panarchy have evolved, to what type of problems they have been applied and found useful, and finally, what key research frontiers remain. Our topic modeling results helped
in identifying some of the problems that the community is addressing and how
they have changed over time. Some of these problems include climate change
(topic 24), urban development and disaster response (7), conservation and
landscape management trade-offs (5), natural resource management (9), food
production (3), policy tools and plans (16), community based management (13) or
knowledge and local social practices (25). We also observed topics closely
related to methods or approaches to study social-ecological systems such as
modeling (19,21), networks (6), or ecosystem services (15). Topic modeling can
be used to further investigate particular academic communities, identify key
landmark papers in certain topics, or identify the network of key collaborators
and experts.
